---
title: "Neural Network with Weather Variables"
author: "Gilman Salzberg"
date: "2026-04-15"
output: pdf_document
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=80), tidy=FALSE,
                      message=FALSE, warning=FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
rm(list = ls())

library(lubridate)
library(ggplot2)
library(forecast)
library(tidyverse)
library(kableExtra)
library(here)

```

## Load Processed Data

```{r}
here()
demand <- read.csv(
  file = "Data/Processed/CleanDemand_v1.csv"
) %>%
  select(-X) %>%
  mutate(date = ymd(date))

temp <- read.csv(
  file = "Data/Processed/CleanTemp_v1.csv"
) %>%
  select(-X) %>%
  mutate(date = ymd(date))

humidity <- read.csv(
  file = "Data/Processed/CleanHumidity_v1.csv"
) %>%
  select(-X) %>%
  mutate(date = ymd(date))

# Join all datasets
data_full <- demand %>%
  left_join(temp, by = "date") %>%
  left_join(humidity, by = "date")

nobs <- nrow(data_full)

head(data_full)
summary(data_full)
```

## Create msts Object and Weather Matrix

```{r}
# Handle NAs BEFORE creating msts (7 NAs from DST transitions)
demand_clean <- na.interp(data_full$avg_load, linear = TRUE)

# Create msts from clean data
demand_msts <- msts(demand_clean,
                    seasonal.periods = c(7, 365.25),
                    start = c(2005, 1))

cat("NAs remaining:", sum(is.na(demand_msts)), "\n")
cat("demand_msts class:", class(demand_msts), "\n")

# Create matrix of exogenous variables (temperature, humidity, and fourier terms)
weather_xreg <- cbind(temp = data_full$avg_temp,
                      humid = data_full$avg_humidity)

head(weather_xreg)
```

## Training / Test Split

Train on Jan 1 2005 - June 30 2010 and test on July 1 2010 - June 30 2011

```{r}

n_for <- 365

# DON'T use subset() - it loses msts structure
# Recreate msts objects from actual data

# Training data - interpolate BEFORE creating msts
train_load <- data_full$avg_load[1:(nobs - n_for)]
train_load_clean <- na.interp(train_load, linear = TRUE)
demand_train <- msts(train_load_clean,
                     seasonal.periods = c(7, 365.25),
                     start = c(2005, 1))

# Test data
test_load <- data_full$avg_load[(nobs - n_for + 1):nobs]
test_load_clean <- na.interp(test_load, linear = TRUE)
demand_test <- msts(test_load_clean,
                    seasonal.periods = c(7, 365.25),
                    start = c(2010, 182))

# Split weather variables
weather_train <- weather_xreg[1:(nobs - n_for), ]
weather_test <- weather_xreg[(nobs - n_for + 1):nobs, ]

# For July 2010 validation (first 31 days of test)
weather_july2010 <- weather_test[1:31, ]
observed_july2010 <- data_full$avg_load[(nobs - n_for + 1):(nobs - n_for + 31)]

# Verify msts structure preserved
cat("demand_train class:", class(demand_train), "\n")

autoplot(demand_train)
autoplot(demand_test)
```

## Fit Neural Network with Fourier Terms and Weather

Following class example from Module 9.
NNETAR uses lagged values and can include xreg for exogenous variables.
We'll use Fourier terms K=c(2,4) plus weather variables.

```{r}
# Neural Network with Fourier terms and weather variables
# p = number of non-seasonal lags
# P = number of seasonal lags (0 since we use Fourier for seasonality)
# xreg = exogenous regressors (Fourier terms + weather)

NN_weather_fit <- nnetar(demand_train,
                        p = 1,
                        P = 0,
                        xreg = cbind(fourier(demand_train, K = c(2, 4)),
                                    weather_train))

print(NN_weather_fit)
```

## Forecast for Validation (July 2010)

```{r}
# Forecast with Neural Network
# Need to provide fourier terms AND weather data for forecast period
NN_weather_for <- forecast(NN_weather_fit,
                           h = 31,
                           xreg = cbind(fourier(demand_train, K = c(2, 4), h = 31),
                                       weather_july2010))

# Plot forecasting results
autoplot(NN_weather_for) +
  ylab("Average Daily Load (kWh)") +
  ggtitle("Neural Network + Weather: Forecast")
```

## Plot Forecast vs Observed for July 2010

```{r}
autoplot(demand_test) +
  autolayer(NN_weather_for, series="NN+Weather", PI=FALSE) +
  coord_cartesian(xlim = c(2010.5, 2010.5 + 31/365.25)) +
  ylab("Average Daily Load (kWh)") +
  xlab("Date") +
  labs(title = "Neural Network+Weather: Forecast vs. Observed - July 2010",
       color = "") +
  theme_classic()
```

## Check Accuracy

```{r}
NN_weather_scores <- accuracy(NN_weather_for$mean, observed_july2010)

scores_df <- as.data.frame(NN_weather_scores)
row.names(scores_df) <- c("NN+Weather")

kbl(scores_df,
    caption = "Neural Network+Weather Forecast Accuracy - July 2010 Validation",
    digits = array(4, ncol(scores_df))) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = "hold_position")
```

## Final Forecast for July 2011

Retrain on the full dataset (Jan 2005 - Jun 2011) and forecast July 2011.

**NOTE**: Using July 2010 weather as proxy for July 2011 (reasonable seasonal assumption).

```{r}
# Extract July 2010 weather
july2010_weather <- weather_xreg[(nobs - n_for + 1):(nobs - n_for + 31), ]

# Retrain Neural Network on full data
NN_weather_final <- nnetar(demand_msts,
                          p = 1,
                          P = 0,
                          xreg = cbind(fourier(demand_msts, K = c(2, 4)),
                                      weather_xreg))

# Forecast July 2011
NN_weather_final_for <- forecast(NN_weather_final,
                                h = 31,
                                xreg = cbind(fourier(demand_msts, K = c(2, 4), h = 31),
                                            july2010_weather))

autoplot(NN_weather_final_for) +
  ylab("Average Daily Load (kWh)") +
  xlab("Year") +
  ggtitle("Neural Network+Weather: July 2011 Forecast") +
  theme_classic()
```

## Export Forecast for Kaggle

```{r}
july2011_dates <- seq(as.Date("2011-07-01"), as.Date("2011-07-31"), by = "day")

submission <- data.frame(
  date = july2011_dates,
  load = as.numeric(NN_weather_final_for$mean)
)

write.csv(submission,
          file = "Submissions/gilman_salzberg_nn_weather.csv",
          row.names = FALSE)

head(submission)
```
