---
title: "Seasonal Naive Model 2"
date: "2026-04-02"
output: pdf_document
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=80), tidy=FALSE,
                      message=FALSE, warning=FALSE)
```

```{r packages, message=FALSE, warning=FALSE}
rm(list = ls())

library(lubridate)
library(ggplot2)
library(forecast)
library(tidyverse)
library(kableExtra)
```

## Load Processed Data

```{r}
here()
demand <- read.csv(
  file = "Data/Processed/CleanDemand_v1.csv"
) %>%
  select(-X) %>%
  mutate(date = ymd(date))

nobs <- nrow(demand)

head(demand)
summary(demand)
```

## Create msts Object

Using msts() to capture both the weekly (period = 7) and the annual cycle (period = 365.25)

```{r}
demand_msts <- msts(demand$avg_load,
                    seasonal.periods = c(7, 365.25),
                    start = c(2005, 1))
```

## Visualize and Decompose

```{r}
autoplot(demand_msts) +
  ylab("Average Daily Load (kWh)") +
  xlab("Year") +
  ggtitle("Daily Electricity Demand: Jan 2005 - Jun 2011")
```

```{r}

demand_msts %>%
  mstl() %>%
  autoplot()
```

## Training / Test Split

Train on Jan 1 2005 - June 30 2010 and test on July 1 2010 - June 31 2011

```{r}

n_for <- 365

demand_train <- subset(demand_msts,
                       end = length(demand_msts) - n_for)

demand_test <- subset(demand_msts,
                      start = length(demand_msts) - n_for + 1)

# validation is July 2010 only - first 31 days of test window
observed_july2010 <- demand$avg_load[(nobs - n_for + 1):(nobs - n_for + 31)]

autoplot(demand_train)
autoplot(demand_test)
```

## Fit Seasonal Naive Model

```{r}
SNAIVE_for <- snaive(demand_train, h = 31)

# Residual diagnostics
checkresiduals(SNAIVE_for)
```

## Check accuracy

```{r}
SNAIVE_scores <- accuracy(SNAIVE_for$mean, observed_july2010)

scores_df <- as.data.frame(SNAIVE_scores)
row.names(scores_df) <- c("SNAIVE")

kbl(scores_df,
    caption = "Seasonal Naive Forecast Accuracy - July 2010 Validation",
    digits = array(4, ncol(scores_df))) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = "hold_position")
```

## Plot forecast vs observed for July 2010

```{r}

autoplot(demand_test) +
  autolayer(SNAIVE_for, PI=FALSE, series="Seasonal Naive") +
  coord_cartesian(xlim = c(2010.5, 2010.5 + 31/365.25)) +
  ylab("Average Daily Load (kWh)") +
  xlab("Date") +
  labs(title = "Seasonal Naive: Forecast vs. Observed - July 2010",
       color = "") +
  theme_classic()
```

## Final forecast

Retrain on the full dataset (Jan 2005 - Jun 2011) and forecast July 2011

```{r}

SNAIVE_final <- snaive(demand_msts, h = 31)
autoplot(SNAIVE_final) +
  ylab("Average Daily Load (kWh)") +
  xlab("Year") +
  ggtitle("Seasonal Naive: July 2011 Forecast") +
  theme_classic()
```

## Export forecast for Kaggle

```{r}
july2011_dates <- seq(as.Date("2011-07-01"), as.Date("2011-07-31"), by = "day")

submission <- data.frame(
  date = july2011_dates,
  load = as.numeric(SNAIVE_final$mean)
)

write.csv(submission,
          file = "Submissions/gilman_salzberg_snaive2.csv",
          row.names = FALSE)

head(submission)
```
