---
title: "STL+ETS Model"
author: "Gilman Salzberg"
date: "2026-04-21"
output: pdf_document
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r packages}
rm(list = ls())

library(lubridate)
library(ggplot2)
library(forecast)
library(tidyverse)
library(kableExtra)
library(here)
```

## Load Processed Data

```{r}
here()
demand <- read.csv("Data/Processed/CleanDemand_v1.csv") %>%
  select(-X) %>%
  mutate(date = ymd(date))

nobs <- nrow(demand)

head(demand)
summary(demand)
```

## Create msts Object

```{r}
# Handle NAs before creating msts
demand_clean <- na.interp(demand$avg_load, linear = TRUE)

demand_msts <- msts(demand_clean,
                    seasonal.periods = c(7, 365.25),
                    start = c(2005, 1))

cat("NAs remaining:", sum(is.na(demand_msts)), "\n")
```

## Visualize Series

```{r}
autoplot(demand_msts) +
  ylab("Average Daily Load (kWh)") +
  xlab("Year") +
  ggtitle("Daily Electricity Demand: Jan 2005 - Jun 2011")
```

## Training / Test Split

```{r}
n_for <- 365

# Recreate msts for training (don't use subset - loses msts structure)
train_load <- demand$avg_load[1:(nobs - n_for)]
train_load_clean <- na.interp(train_load, linear = TRUE)
demand_train <- msts(train_load_clean,
                     seasonal.periods = c(7, 365.25),
                     start = c(2005, 1))

# Test data
test_load <- demand$avg_load[(nobs - n_for + 1):nobs]
test_load_clean <- na.interp(test_load, linear = TRUE)
demand_test <- msts(test_load_clean,
                    seasonal.periods = c(7, 365.25),
                    start = c(2010, 182))

# Validation observations
observed_july2010 <- demand$avg_load[(nobs - n_for + 1):(nobs - n_for + 31)]

autoplot(demand_train)
autoplot(demand_test)
```

## Fit STL+ETS Model

STL = Seasonal-Trend decomposition using Loess
ETS = Error-Trend-Seasonal exponential smoothing

This is a 2-step hybrid approach:
1. STL decomposes the series into trend, seasonal, and remainder
2. ETS forecasts the seasonally-adjusted series
3. Seasonal component is added back

```{r}
cat("Fitting STL+ETS model...\n")

# STL + ETS with periodic seasonal window
ETS_fit <- stlf(demand_train, 
                h = 31,
                s.window = "periodic")

print(ETS_fit)

# Check the ETS model selected
cat("\nETS model selected:", ETS_fit$model$method, "\n")
```

## Check Residuals

```{r}
checkresiduals(ETS_fit)
```

## Forecast Plots

```{r}
# Plot forecasting results
autoplot(ETS_fit) +
  ylab("Average Daily Load (kWh)") +
  ggtitle("STL+ETS Forecast")

# Plot model + observed data for validation
autoplot(demand_test) +
  autolayer(ETS_fit, series="STL+ETS", PI=FALSE) +
  coord_cartesian(xlim = c(2010.5, 2010.5 + 31/365.25)) +
  ylab("Average Daily Load (kWh)") +
  xlab("Date") +
  labs(title = "STL+ETS: Forecast vs. Observed - July 2010",
       color = "") +
  theme_classic()
```

## Check Accuracy

```{r}
ETS_scores <- accuracy(ETS_fit$mean, observed_july2010)

scores_df <- as.data.frame(ETS_scores)
row.names(scores_df) <- c("STL+ETS")

kbl(scores_df,
    caption = "STL+ETS Forecast Accuracy - July 2010 Validation",
    digits = array(4, ncol(scores_df))) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = "hold_position")

cat("\nSTL+ETS Validation Performance:\n")
cat("RMSE:", round(ETS_scores[2], 2), "\n")
cat("MAPE:", round(ETS_scores[5], 2), "%\n")
```

## Try Alternative STL+ETS Configurations

```{r}
cat("\n=== TESTING ALTERNATIVE CONFIGURATIONS ===\n\n")

# Configuration 1: Robust STL (resistant to outliers)
cat("1. Robust STL...\n")
ETS_robust <- stlf(demand_train, 
                   h = 31,
                   s.window = "periodic",
                   robust = TRUE)
acc_robust <- accuracy(ETS_robust$mean, observed_july2010)

# Configuration 2: Different seasonal window
cat("2. STL with s.window=13...\n")
ETS_s13 <- stlf(demand_train, 
                h = 31,
                s.window = 13)
acc_s13 <- accuracy(ETS_s13$mean, observed_july2010)

# Configuration 3: Force specific ETS model (damped trend)
cat("3. STL + ETS(A,Ad,A)...\n")
ETS_damped <- stlf(demand_train,
                   h = 31,
                   s.window = "periodic",
                   etsmodel = "ZZZ",  # Auto-select
                   damped = TRUE)
acc_damped <- accuracy(ETS_damped$mean, observed_july2010)

# Compare all configurations
config_results <- data.frame(
  Configuration = c("Standard", "Robust", "s.window=13", "Damped Trend"),
  RMSE = c(ETS_scores[2], acc_robust[2], acc_s13[2], acc_damped[2]),
  MAE = c(ETS_scores[3], acc_robust[3], acc_s13[3], acc_damped[3]),
  MAPE = c(ETS_scores[5], acc_robust[5], acc_s13[5], acc_damped[5])
) %>%
  arrange(RMSE) %>%
  mutate(Rank = row_number())

kbl(config_results,
    caption = "STL+ETS Configuration Comparison",
    digits = 4) %>%
  kable_styling(full_width = FALSE, position = "center",
                latex_options = "hold_position") %>%
  row_spec(1, bold = TRUE, background = "#90EE90")

cat("\nBest configuration:", config_results$Configuration[1], "\n")
```

## Final Forecast for July 2011

Retrain on full dataset and forecast July 2011

```{r}
cat("\n=== GENERATING FINAL FORECASTS ===\n\n")

# Refit all configurations on full data
cat("Refitting models on full dataset...\n")

ETS_final_standard <- stlf(demand_msts, h = 31, s.window = "periodic")

ETS_final_robust <- stlf(demand_msts, h = 31, s.window = "periodic", robust = TRUE)

ETS_final_s13 <- stlf(demand_msts, h = 31, s.window = 13)

ETS_final_damped <- stlf(demand_msts, h = 31, s.window = "periodic", 
                         etsmodel = "ZZZ", damped = TRUE)

# Plot final forecast
autoplot(ETS_final_standard) +
  ylab("Average Daily Load (kWh)") +
  xlab("Year") +
  ggtitle("STL+ETS: July 2011 Forecast (Standard)") +
  theme_classic()
```

## Export Forecasts for Kaggle

```{r}
july2011_dates <- seq(as.Date("2011-07-01"), as.Date("2011-07-31"), by = "day")

# Save all 4 configurations
configs <- list(
  list(name = "standard", model = ETS_final_standard),
  list(name = "robust", model = ETS_final_robust),
  list(name = "s13", model = ETS_final_s13),
  list(name = "damped", model = ETS_final_damped)
)

for(config in configs) {
  submission <- data.frame(
    date = july2011_dates,
    load = as.numeric(config$model$mean)
  )
  
  filename <- paste0("Submissions/STL_ETS/gilman_salzberg_ets_", config$name, ".csv")
  write.csv(submission, file = filename, row.names = FALSE)
  cat("Saved:", filename, "\n")
}

cat("\n=== COMPLETE ===\n")
cat("Generated 4 STL+ETS submission files\n")
cat("Submit the best performing configuration based on validation!\n")
```

## Summary

```{r}
cat("\n=== STL+ETS SUMMARY ===\n\n")
cat("Best Configuration:", config_results$Configuration[1], "\n")
cat("Validation RMSE:", round(config_results$RMSE[1], 2), "\n")
cat("Validation MAPE:", round(config_results$MAPE[1], 2), "%\n\n")

cat("STL+ETS is fast, stable, and often competitive!\n")
cat("It handles trend and seasonality automatically.\n")
cat("Submit to Kaggle and compare with your other models.\n")
```
