---
title: "Kaggle Submission Analysis"
author: "Gilman Salzberg"
date: "2026-04-21"
output: pdf_document
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8)
```

```{r packages}
library(tidyverse)
library(here)
library(lubridate)

here()
```

## Load All Kaggle Submissions

```{r}
# Get all CSV files in Submissions folder (including subfolders)
submission_files <- list.files("Submissions", 
                              pattern = "*.csv", 
                              full.names = TRUE, 
                              recursive = TRUE)

# Exclude the template file
submission_files <- submission_files[!grepl("template", submission_files)]

cat("Found", length(submission_files), "submission files\n\n")

# Load all submissions into a list
submissions_list <- list()

for(file in submission_files) {
  # Extract model name from filename
  model_name <- gsub("gilman_salzberg_", "", basename(file))
  model_name <- gsub(".csv", "", model_name)
  
  # Load data
  data <- read.csv(file)
  data$model <- model_name
  data$date <- ymd(data$date)
  
  submissions_list[[model_name]] <- data
  
  cat("Loaded:", model_name, "- Mean forecast:", round(mean(data$load), 0), "\n")
}

# Combine all submissions
all_submissions <- bind_rows(submissions_list)

# Remove models with NA dates
all_submissions <- all_submissions %>%
  filter(!is.na(date))

cat("Removed rows with NA dates\n")
cat("\nTotal submissions loaded:", length(unique(all_submissions$model)), "\n")


```

## Your Known Kaggle Scores

Manually enter the scores you've seen on Kaggle:

```{r}

kaggle_scores <- data.frame(
  model = c("bayesian_k100", "super_ensemble", "advanced_bayesian_v2", 
            "adaptive_critical", "bayesian_top5",
            "optimized_trimmed", "advanced_winsorized", 
            "ULTIMATE_v2", "ensemble_top3", "optimized_diverse",
            "top3_kaggle_ensemblev2", "advanced_trimmed25", "advanced_trimmed20_v2",
            "optimized_median", "optimized_top3_weighted", "optimized_smoothed", 
            "top3_kaggle_ensemble", "ensemble_simple", "ensemble_weighted",
            "arima_fourier_weather", "advanced_stacked_v2", "arima_weather_K2_8", 
            "arima_weather_K3_8", "arima_weather_K3_6", "arima_weather_K2_6", 
            "ensemble_arima_only", "tbats", "ensemble_best2", "ets_s13", 
            "arima_weather_K1_6", "advanced_hybrid_v2", "nn_weather", 
            "ets_robust", "arima_fourier", "snaive2", "ets_damped", "ets_standard"),
  score = c(0.1001, 0.1001, 0.1003, 0.1003, 0.1003,
            0.1008, 0.1013, 0.1014, 0.1015, 0.1015, 0.1017, 0.1018, 0.1018,
            0.1019, 0.1019, 0.1022, 0.1023, 0.1025, 0.1037, 0.1039, 0.1040,
            0.1060, 0.1062, 0.1078, 0.1079, 0.1082, 0.1110, 0.1121, 0.1318,
            0.1321, 0.1323, 0.1331, 0.1465, 0.1528, 0.1602, 0.1647, 0.1648),
  stringsAsFactors = FALSE
)

# Remove duplicates (keep best score for each model)
kaggle_scores <- kaggle_scores %>%
  group_by(model) %>%
  slice_min(score, n = 1) %>%
  ungroup()

# Sort by score
kaggle_scores <- kaggle_scores %>%
  arrange(score) %>%
  mutate(rank = row_number())

cat("Total models with Kaggle scores:", nrow(kaggle_scores), "\n")
cat("Best score:", min(kaggle_scores$score), "\n")
cat("Worst score:", max(kaggle_scores$score), "\n")
print(kaggle_scores)
```

## Summary Statistics for Each Submission

```{r}
submission_stats <- all_submissions %>%
  group_by(model) %>%
  summarize(
    mean_forecast = mean(load),
    min_forecast = min(load),
    max_forecast = max(load),
    sd_forecast = sd(load),
    range = max_forecast - min_forecast
  ) %>%
  left_join(kaggle_scores, by = "model") %>%
  arrange(score)

knitr::kable(submission_stats, digits = 2,
             caption = "Forecast Statistics by Model (Ranked by Kaggle Score)")
```

## Plot All Submissions Together

```{r fig.height=10, fig.width=14}
# Color code by performance
all_submissions_scored <- all_submissions %>%
  left_join(kaggle_scores %>% select(model, score), by = "model")

# Plot all forecasts
ggplot(all_submissions_scored, aes(x = date, y = load, color = model, group = model)) +
  geom_line(alpha = 0.7, size = 0.8) +
  scale_x_date(date_breaks = "3 days", date_labels = "%b %d") +
  labs(title = "All Kaggle Submissions - July 2011 Forecasts",
       subtitle = paste(length(unique(all_submissions$model)), "different models"),
       x = "Date",
       y = "Forecasted Load (kWh)",
       color = "Model") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))

```

## Plot Top 5 vs Bottom 5 Performers

```{r fig.height=8}
top5_models <- kaggle_scores %>% slice(1:5) %>% pull(model)
bottom5_models <- kaggle_scores %>% slice((n()-4):n()) %>% pull(model)

# Top 5
top5_data <- all_submissions %>%
  filter(model %in% top5_models) %>%
  left_join(kaggle_scores, by = "model")

ggplot(top5_data, aes(x = date, y = load, color = reorder(model, score), group = model)) +
  geom_line(size = 1.2) +
  scale_x_date(date_breaks = "3 days", date_labels = "%b %d") +
  labs(title = "Top 5 Best Performing Models on Kaggle",
       x = "Date",
       y = "Forecasted Load (kWh)",
       color = "Model (by score)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bottom 5
bottom5_data <- all_submissions %>%
  filter(model %in% bottom5_models) %>%
  left_join(kaggle_scores, by = "model")

ggplot(bottom5_data, aes(x = date, y = load, color = reorder(model, -score), group = model)) +
  geom_line(size = 1.2) +
  scale_x_date(date_breaks = "3 days", date_labels = "%b %d") +
  labs(title = "Bottom 5 Worst Performing Models on Kaggle",
       x = "Date",
       y = "Forecasted Load (kWh)",
       color = "Model (by score)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

## Identify Common Patterns

```{r}

# Calculate correlation between forecasts
forecast_wide <- all_submissions %>%
  select(date, model, load) %>%
  pivot_wider(names_from = model, values_from = load)

# Unlist all columns except date
for(col in names(forecast_wide)[-1]) {
  forecast_wide[[col]] <- unlist(forecast_wide[[col]])
}

# Remove date column for correlation
numeric_cols <- forecast_wide %>% select(-date)

# Calculate correlation
cor_matrix <- cor(numeric_cols, use = "pairwise.complete.obs")

# Find which models are most similar
cat("\n=== MOST CORRELATED MODELS ===\n")
cor_df <- as.data.frame(as.table(cor_matrix))
names(cor_df) <- c("Model1", "Model2", "Correlation")
cor_df <- cor_df %>%
  filter(Model1 != Model2, Correlation < 1) %>%
  arrange(desc(Correlation)) %>%
  distinct(Correlation, .keep_all = TRUE) %>%
  slice(1:10)

print(cor_df)

```



## Day-by-Day Variation Analysis

Which days are hardest to predict?

```{r fig.height=6}
# Use only good models (score < 0.11)
good_model_names <- kaggle_scores %>% filter(score < 0.11) %>% pull(model)

# Calculate range of predictions for each day (WINNERS ONLY)
day_variation <- all_submissions %>%
  filter(model %in% good_model_names) %>%
  group_by(date) %>%
  summarize(
    min_pred = min(load),
    max_pred = max(load),
    mean_pred = mean(load),
    range = max_pred - min_pred,
    cv = sd(load) / mean(load) * 100  # Coefficient of variation
  )

ggplot(day_variation, aes(x = date)) +
  geom_ribbon(aes(ymin = min_pred, ymax = max_pred), alpha = 0.3, fill = "blue") +
  geom_line(aes(y = mean_pred), color = "red", size = 1.2) +
  scale_x_date(date_breaks = "3 days", date_labels = "%b %d") +
  labs(title = "Range of TOP Model Predictions by Day (Score < 0.11)",
       subtitle = "Red line = mean prediction, Blue band = min to max range",
       x = "Date",
       y = "Load (kWh)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Which days have highest disagreement AMONG WINNERS?
cat("\n=== DAYS WITH HIGHEST DISAGREEMENT (TOP MODELS ONLY) ===\n")
print(day_variation %>% arrange(desc(range)) %>% select(date, range, cv) %>% head(10))
```

## Best Model Analysis

```{r}
best_model <- kaggle_scores$model[1]
best_forecast <- all_submissions %>% filter(model == best_model)

cat("\n=== BEST MODEL: ", best_model, "===\n")
cat("Kaggle Score:", kaggle_scores$score[1], "\n\n")

cat("Forecast Summary:\n")
summary(best_forecast$load)

# Compare to ensemble average
ensemble_avg <- all_submissions %>%
  group_by(date) %>%
  summarize(avg_load = mean(load))

comparison <- best_forecast %>%
  left_join(ensemble_avg, by = "date") %>%
  mutate(diff = load - avg_load)

cat("\nHow different from ensemble average?\n")
cat("Mean difference:", round(mean(comparison$diff), 2), "\n")
cat("Max difference:", round(max(abs(comparison$diff)), 2), "\n")

ggplot(comparison, aes(x = date)) +
  geom_line(aes(y = load, color = "Best Model"), size = 1.2) +
  geom_line(aes(y = avg_load, color = "Ensemble Avg"), size = 1.2, linetype = "dashed") +
  scale_x_date(date_breaks = "3 days", date_labels = "%b %d") +
  labs(title = paste("Best Model (", best_model, ") vs Ensemble Average"),
       x = "Date",
       y = "Load (kWh)",
       color = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Recommendations Based on Analysis

```{r}
cat("\n=== ANALYSIS INSIGHTS ===\n\n")

# 1. Check if best models are conservative or aggressive
best3_avg <- all_submissions %>%
  filter(model %in% top5_models[1:3]) %>%
  summarize(mean_load = mean(load)) %>%
  pull(mean_load)

overall_avg <- mean(all_submissions$load)

if(best3_avg < overall_avg) {
  cat("1. Best models predict LOWER than average\n")
  cat("   → Try models that predict conservatively\n\n")
} else {
  cat("1. Best models predict HIGHER than average\n")
  cat("   → Try more aggressive forecasts\n\n")
}

# 2. Check variability
best3_sd <- all_submissions %>%
  filter(model %in% top5_models[1:3]) %>%
  group_by(model) %>%
  summarize(sd = sd(load)) %>%
  summarize(mean_sd = mean(sd)) %>%
  pull(mean_sd)

overall_sd <- all_submissions %>%
  group_by(model) %>%
  summarize(sd = sd(load)) %>%
  summarize(mean_sd = mean(sd)) %>%
  pull(mean_sd)

if(best3_sd < overall_sd) {
  cat("2. Best models are SMOOTHER (less variable)\n")
  cat("   → Try removing high-frequency noise\n\n")
} else {
  cat("2. Best models are MORE VARIABLE\n")
  cat("   → Try capturing more day-to-day variation\n\n")
}

# 3. Model diversity
cat("3. You have", length(unique(all_submissions$model)), "submissions\n")
cat("   Top teams have 20-24 submissions\n")
cat("   → Keep iterating! More submissions = more chances\n\n")

cat("4. Try creating a custom ensemble of your top 3 Kaggle performers\n")
cat("   Models:", paste(top5_models[1:3], collapse = ", "), "\n")
```

## Generate New Ensemble from Top Performers

```{r}
cat("\n=== CREATING NEW ENSEMBLE FROM TOP 3 KAGGLE PERFORMERS ===\n\n")

top3_ensemble <- all_submissions %>%
  filter(model %in% top5_models[1:3]) %>%
  group_by(date) %>%
  summarize(load = mean(load))

# Save submission
submission <- data.frame(
  date = top3_ensemble$date,
  load = top3_ensemble$load
)

write.csv(submission,
          file = "Submissions/ensemble_models/gilman_salzberg_top3_kaggle_ensemblev2.csv",
          row.names = FALSE)

cat("Created: gilman_salzberg_top3_kaggle_ensemble.csv\n")
cat("This averages your 3 best Kaggle performers\n")
cat("\nForecast summary:\n")
summary(submission$load)
```
