---
title: "Ultimate Bayesian Fine-Tuning - Final Push to 0.100"
author: "Gilman Salzberg"
date: "2026-04-21"
output: pdf_document
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r packages}
library(tidyverse)
library(here)
library(lubridate)

here()
```

## Load All Submissions

```{r}
submission_files <- list.files("Submissions", pattern = "*.csv", 
                               full.names = TRUE, recursive = TRUE)
submission_files <- submission_files[!grepl("template", submission_files)]

submissions_list <- list()

for(file in submission_files) {
  model_name <- gsub("gilman_salzberg_", "", basename(file))
  model_name <- gsub(".csv", "", model_name)
  
  data <- read.csv(file)
  data$date <- suppressWarnings(ymd(data$date))
  
  if(sum(!is.na(data$date)) > 0) {
    data$model <- model_name
    submissions_list[[model_name]] <- data
  }
}

all_submissions <- bind_rows(submissions_list) %>%
  filter(!is.na(date))

cat("Loaded", length(unique(all_submissions$model)), "models\n")
```

## Kaggle Scores

```{r}
kaggle_scores <- data.frame(
  model = c("advanced_bayesian_v2", "optimized_trimmed", "advanced_winsorized", 
            "ULTIMATE_v2", "ensemble_top3", "optimized_diverse",
            "top3_kaggle_ensemblev2", "advanced_trimmed25", "advanced_trimmed20_v2",
            "optimized_median", "optimized_top3_weighted", "optimized_smoothed", 
            "top3_kaggle_ensemble", "ensemble_simple", "ensemble_weighted",
            "arima_fourier_weather", "advanced_stacked_v2", "arima_weather_K2_8", 
            "arima_weather_K3_8", "arima_weather_K3_6", "arima_weather_K2_6", 
            "ensemble_arima_only", "tbats", "ensemble_best2", "ets_s13", 
            "arima_weather_K1_6", "advanced_hybrid_v2", "nn_weather", 
            "ets_robust", "arima_fourier", "snaive2", "ets_damped", "ets_standard"),
  score = c(0.1003, 0.1008, 0.1013, 0.1014, 0.1015, 0.1015, 0.1017, 0.1018, 0.1018,
            0.1019, 0.1019, 0.1022, 0.1023, 0.1025, 0.1037, 0.1039, 0.1040,
            0.1060, 0.1062, 0.1078, 0.1079, 0.1082, 0.1110, 0.1121, 0.1318,
            0.1321, 0.1323, 0.1331, 0.1465, 0.1528, 0.1602, 0.1647, 0.1648),
  stringsAsFactors = FALSE
)

kaggle_scores <- kaggle_scores %>%
  group_by(model) %>%
  slice_min(score, n = 1) %>%
  ungroup() %>%
  arrange(score)

cat("Using models with scores\n")
```

## STRATEGY 1: Hyper-Aggressive Bayesian (k=30, 40, 50, 60)

Test multiple aggression levels to maximize weight on the best model:

```{r}
cat("\n=== STRATEGY 1: HYPER-AGGRESSIVE BAYESIAN ===\n\n")

good_models <- kaggle_scores %>% filter(score < 0.115) %>% pull(model)

# Test different k values (higher = more aggressive)
k_values <- c(20, 30, 40, 50, 60, 80, 100)

bayesian_variants <- list()

for(k_val in k_values) {
  
  cat("Testing k =", k_val, "...\n")
  
  # Calculate weights
  weights <- kaggle_scores %>%
    filter(model %in% good_models) %>%
    mutate(
      raw_weight = exp(-k_val * score),
      normalized_weight = raw_weight / sum(raw_weight)
    ) %>%
    arrange(desc(normalized_weight))
  
  # Show top 5 weights
  cat("  Top 5 weights:\n")
  print(weights %>% select(model, score, normalized_weight) %>% head(5))
  
  # Create ensemble
  ensemble <- all_submissions %>%
    filter(model %in% weights$model) %>%
    left_join(weights %>% select(model, normalized_weight), by = "model") %>%
    group_by(date) %>%
    summarize(load = sum(load * normalized_weight), .groups = "drop")
  
  bayesian_variants[[paste0("k", k_val)]] <- ensemble
  
  cat("  Mean:", round(mean(ensemble$load), 2), 
      "SD:", round(sd(ensemble$load), 2), "\n\n")
}
```

## STRATEGY 2: Bayesian Top-N Only (Reduce Noise)

Use Bayesian weighting but ONLY on the top 3, 5, 7, or 10 models:

```{r}
cat("\n=== STRATEGY 2: BAYESIAN TOP-N ONLY ===\n\n")

top_n_values <- c(3, 5, 7, 10)
k_optimal <- 40  # Use aggressive weighting

bayesian_topn <- list()

for(n_val in top_n_values) {
  
  cat("Testing Top", n_val, "models only...\n")
  
  # Get top N models
  top_models <- kaggle_scores %>% 
    arrange(score) %>% 
    slice(1:n_val) %>% 
    pull(model)
  
  # Calculate weights
  weights <- kaggle_scores %>%
    filter(model %in% top_models) %>%
    mutate(
      raw_weight = exp(-k_optimal * score),
      normalized_weight = raw_weight / sum(raw_weight)
    )
  
  cat("  Weights:\n")
  print(weights %>% select(model, score, normalized_weight))
  
  # Create ensemble
  ensemble <- all_submissions %>%
    filter(model %in% weights$model) %>%
    left_join(weights %>% select(model, normalized_weight), by = "model") %>%
    group_by(date) %>%
    summarize(load = sum(load * normalized_weight), .groups = "drop")
  
  bayesian_topn[[paste0("top", n_val)]] <- ensemble
  
  cat("  Mean:", round(mean(ensemble$load), 2), 
      "SD:", round(sd(ensemble$load), 2), "\n\n")
}
```

## STRATEGY 3: Critical Date Adaptive Bayesian

Give bonus weight to models that perform well on the hardest dates:

```{r}
cat("\n=== STRATEGY 3: CRITICAL DATE ADAPTIVE BAYESIAN ===\n\n")

# Define critical dates (highest disagreement among top models)
critical_dates <- as.Date(c("2011-07-03", "2011-07-31", "2011-07-02", 
                            "2011-07-25", "2011-07-04"))

cat("Critical dates:", paste(critical_dates, collapse = ", "), "\n\n")

# Calculate how well each model handles critical dates
# (compare to median of top 5 models on those dates)
top5_models <- kaggle_scores %>% slice(1:5) %>% pull(model)

critical_performance <- all_submissions %>%
  filter(date %in% critical_dates, model %in% good_models) %>%
  group_by(date) %>%
  mutate(
    top5_median = median(load[model %in% top5_models])
  ) %>%
  ungroup() %>%
  group_by(model) %>%
  summarize(
    critical_error = mean(abs(load - top5_median)),
    .groups = "drop"
  )

cat("Critical date performance (lower error = better):\n")
print(critical_performance %>% arrange(critical_error) %>% head(10))

# Combine Kaggle score + critical date performance
combined_scores <- kaggle_scores %>%
  filter(model %in% good_models) %>%
  left_join(critical_performance, by = "model") %>%
  mutate(
    # Normalize both scores to 0-1 range
    kaggle_norm = (score - min(score)) / (max(score) - min(score)),
    critical_norm = (critical_error - min(critical_error, na.rm=TRUE)) / 
                    (max(critical_error, na.rm=TRUE) - min(critical_error, na.rm=TRUE)),
    # Combined score (70% Kaggle, 30% critical date performance)
    combined_score = 0.7 * kaggle_norm + 0.3 * critical_norm,
    # Calculate weights
    raw_weight = exp(-40 * combined_score),
    normalized_weight = raw_weight / sum(raw_weight)
  ) %>%
  arrange(desc(normalized_weight))

cat("\nTop 10 by combined score:\n")
print(combined_scores %>% 
      select(model, score, critical_error, combined_score, normalized_weight) %>% 
      head(10))

# Create adaptive ensemble
adaptive_ensemble <- all_submissions %>%
  filter(model %in% combined_scores$model) %>%
  left_join(combined_scores %>% select(model, normalized_weight), by = "model") %>%
  group_by(date) %>%
  summarize(load = sum(load * normalized_weight), .groups = "drop")

cat("\nAdaptive ensemble stats:\n")
cat("Mean:", round(mean(adaptive_ensemble$load), 2), "\n")
cat("SD:", round(sd(adaptive_ensemble$load), 2), "\n")
```

## STRATEGY 4: Date-Specific Switching

Use DIFFERENT model combinations on critical vs regular dates:

```{r}
cat("\n=== STRATEGY 4: DATE-SPECIFIC SWITCHING ===\n\n")

# Prepare weights for regular dates
regular_weights <- kaggle_scores %>%
  filter(model %in% good_models) %>%
  mutate(
    raw_weight = exp(-50 * score),
    normalized_weight = raw_weight / sum(raw_weight)
  )

# Get top 3 critical performers
critical_top3 <- critical_performance %>% 
  arrange(critical_error) %>% 
  slice(1:3) %>% 
  pull(model)

cat("Models for critical dates:", paste(critical_top3, collapse = ", "), "\n")
cat("Using k=50 Bayesian for regular dates\n\n")

# Build ensemble
switching_ensemble <- all_submissions %>%
  mutate(is_critical = date %in% critical_dates) %>%
  left_join(regular_weights %>% select(model, normalized_weight), by = "model") %>%
  group_by(date, is_critical) %>%
  summarize(
    load = if(unique(is_critical)) {
      # Critical date: equal weight to top 3
      mean(load[model %in% critical_top3])
    } else {
      # Regular date: Bayesian weighted
      sum(load * normalized_weight, na.rm = TRUE)
    },
    .groups = "drop"
  )

cat("Switching ensemble stats:\n")
cat("Mean:", round(mean(switching_ensemble$load), 2), "\n")
cat("SD:", round(sd(switching_ensemble$load), 2), "\n")

```

## STRATEGY 5: Super Ensemble (Combine Best Bayesian Variants)

Take the best variants from above and ensemble them:

```{r}
cat("\n=== STRATEGY 5: SUPER ENSEMBLE ===\n\n")

# Combine k=50, top5, and adaptive
super_ensemble <- data.frame(
  date = bayesian_variants$k50$date,
  k50 = bayesian_variants$k50$load,
  top5 = bayesian_topn$top5$load,
  adaptive = adaptive_ensemble$load
) %>%
  mutate(
    # Weighted average: 40% k50, 30% top5, 30% adaptive
    load = 0.4 * k50 + 0.3 * top5 + 0.3 * adaptive
  ) %>%
  select(date, load)

cat("Super ensemble stats:\n")
cat("Mean:", round(mean(super_ensemble$load), 2), "\n")
cat("SD:", round(sd(super_ensemble$load), 2), "\n")
```

## Compare All Strategies

```{r}
cat("\n=== COMPARISON OF ALL STRATEGIES ===\n\n")

comparison <- data.frame(
  Strategy = c(
    paste0("Bayesian k=", k_values),
    paste0("Bayesian Top", top_n_values),
    "Adaptive (70% Kaggle, 30% Critical)",
    "Date-Specific Switching",
    "Super Ensemble"
  ),
  Mean = c(
    sapply(bayesian_variants, function(x) mean(x$load)),
    sapply(bayesian_topn, function(x) mean(x$load)),
    mean(adaptive_ensemble$load),
    mean(switching_ensemble$load),
    mean(super_ensemble$load)
  ),
  SD = c(
    sapply(bayesian_variants, function(x) sd(x$load)),
    sapply(bayesian_topn, function(x) sd(x$load)),
    sd(adaptive_ensemble$load),
    sd(switching_ensemble$load),
    sd(super_ensemble$load)
  )
)

knitr::kable(comparison, digits = 2,
             caption = "All Bayesian Fine-Tuning Strategies")

# Highlight strategies matching winner profile
cat("\n=== STRATEGIES MATCHING WINNER PROFILE ===\n")
cat("(Mean: 4000-4100, SD: 180-220)\n\n")

matching <- comparison %>%
  filter(Mean >= 4000 & Mean <= 4100, SD >= 180 & SD <= 220)

if(nrow(matching) > 0) {
  print(matching)
} else {
  cat("Showing 5 closest to target:\n")
  comparison %>%
    mutate(
      mean_dist = abs(Mean - 4050),
      sd_dist = abs(SD - 200),
      total_dist = mean_dist + sd_dist
    ) %>%
    arrange(total_dist) %>%
    select(Strategy, Mean, SD) %>%
    head(5) %>%
    print()
}
```

## Save Top 10 Strategies for Kaggle

```{r}
july2011_dates <- seq(as.Date("2011-07-01"), as.Date("2011-07-31"), by = "day")

cat("\n=== SAVING TOP 10 STRATEGIES ===\n\n")

# Create list of all strategies
all_strategies <- list(
  list(name = "bayesian_k50", data = bayesian_variants$k50),
  list(name = "bayesian_k60", data = bayesian_variants$k60),
  list(name = "bayesian_k80", data = bayesian_variants$k80),
  list(name = "bayesian_k100", data = bayesian_variants$k100),
  list(name = "bayesian_top3", data = bayesian_topn$top3),
  list(name = "bayesian_top5", data = bayesian_topn$top5),
  list(name = "adaptive_critical", data = adaptive_ensemble),
  list(name = "date_switching", data = switching_ensemble),
  list(name = "super_ensemble", data = super_ensemble),
  list(name = "bayesian_k40_top7", data = bayesian_topn$top7)
)

# Save all
for(strat in all_strategies) {
  submission <- data.frame(
    date = strat$data$date,
    load = as.numeric(strat$data$load)
  )
  
  filename <- paste0("Submissions/ensemble_models/final_push/gilman_salzberg_", 
                    strat$name, ".csv")
  
  # Create directory if needed
  dir.create("Submissions/ensemble_models/final_push", 
             recursive = TRUE, showWarnings = FALSE)
  
  write.csv(submission, file = filename, row.names = FALSE)
  cat("Saved:", filename, "\n")
}

cat("\n=== RECOMMENDATIONS ===\n\n")
cat("Submit these in priority order:\n\n")
cat("1. BAYESIAN_K100 (most aggressive - maximum weight to best model)\n")
cat("   - Expected: 0.0995-0.1005\n\n")
cat("2. SUPER_ENSEMBLE (combines best variants)\n")
cat("   - Expected: 0.0998-0.1006\n\n")
cat("3. BAYESIAN_TOP3 (pure focus on your top 3)\n")
cat("   - Expected: 0.0999-0.1007\n\n")
cat("4. ADAPTIVE_CRITICAL (rewards critical date performance)\n")
cat("   - Expected: 0.1000-0.1008\n\n")
cat("5. DATE_SWITCHING (different strategy for hard dates)\n")
cat("   - Expected: 0.1001-0.1009\n\n")

cat("BEST SHOT: BAYESIAN_K100 or SUPER_ENSEMBLE\n")
cat("Why: Maximum aggression on proven winners + smart combinations\n")
```

## Visualize Weight Distributions

```{r fig.width=10, fig.height=6}
# Compare how weights change with k
weight_comparison <- data.frame()

for(k_val in c(20, 40, 60, 100)) {
  weights <- kaggle_scores %>%
    filter(score < 0.115) %>%
    mutate(
      raw_weight = exp(-k_val * score),
      normalized_weight = raw_weight / sum(raw_weight),
      k = paste0("k=", k_val)
    ) %>%
    select(model, score, k, normalized_weight)
  
  weight_comparison <- bind_rows(weight_comparison, weights)
}

# Plot top 10 models
top10_models <- kaggle_scores %>% slice(1:10) %>% pull(model)

weight_comparison %>%
  filter(model %in% top10_models) %>%
  ggplot(aes(x = reorder(model, score), y = normalized_weight, fill = k)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "How Bayesian Weights Change with Aggression (k)",
       subtitle = "Higher k = more weight to best models",
       x = "Model",
       y = "Normalized Weight",
       fill = "Aggression") +
  theme_minimal()
```
